\LARGE \bf Stochastic Uncertainty Propagation in Power System Dynamics using Measure-valued Proximal Recursions

### 2022-02-27 Papers/Stochastic_Uncertainty_Propagation_TPS/RebuttalRound1-TPWRS-01391-2021.tex

\section*{\large \bf Response to AE:}

\noindent
{\bf AE.1. Comments by AE}:\\
{\em The authors are requested to revise and resubmit their paper after addressing reviewer feedback. Of particular importance is that reviewers point out the model leveraged for analysis is overly simplistic.}

{\nib {\blue We are indebted to the AE and the reviewers for their help to improve the manuscript.

In addition to the itemized responses to the reviewers' comments and the corresponding revisions marked in blue in the manuscript, we have made the \ul{following additional fixes and improvements} (also marked in blue in the revised manuscript):\\
(i) In equation (8), we corrected a typo in its right hand side: the second summand runs over pairs $(i,j)$ such that $i<j$.\\
(ii) To make the presentation tighter, the Appendix appearing in the previous version has been eschewed. We have accordingly adjusted and condensed the material in Sec. III-D of the revised manuscript.\\
(iii) Some mathematical typos have been fixed in the revised manuscript's Sec. III-C and III-D.

@@

\noindent
{\bf R3.2. Comments by reviewer}:\\
{\em The literature is incomplete, many recent works on SDAEs for power system applications are not mentioned.}

{\nib \blue{\ul{As per the reviewer's suggestion, in the revised manuscript, we have included two sentences (marked blue) in Sec. I.1, toward the end of second paragraph, mentioning the usage of SDAEs in power system dynamic simulation}. Therein, the first sentence contains two newly added references: [22] and [23], both on simulating SDAE sample paths.}}

\noindent
{\bf R3.3. Comments by reviewer}:\\
{\em There are several typos in the test (especially in the references).}

{\nib \blue{In Sec. V-A, we corrected ``assuming" $\rightarrow$ ``are assumed". We corrected the two typos pointed out in comment \textbf{R4.8} in this rebuttal document. We also corrected minor mathematical typos in (19), (20a), (22a), as well the transformed domains in Sec. III-D.}}

@@

Following the same vein, there are some unnecessary information and technical points provided in this paper that may be distracting.
For example, on page 3, why would the authors mention the Kuramoto oscillator model?
On page 4, is the detailed discussion regarding Wasserstein-2 metric necessary? Similarly, on page 5, the discussion about the natural energy functional and how the gradient drift case does not apply also seems too detailed.}

{\nib \blue{We have followed the reviewer's suggestion to improve the presentation. \ul{Specifically, in the revised manuscript's Sec. III-A, last two paragraphs summarize the technical relation of this paper w.r.t. the authors' previous work. In particular, we have added three more sentences with a new reference. To improve readability, have also broken what was a single paragraph in the previous version, into two separate paragraphs.}

We appreciate the reviewer's point that the sentence containing equation (6) and the mention of Kuramoto model can be removed without affecting the rest of the development. However, we prefer to keep it since it adds didactic value, and many engineers may feel comfortable by the explicit mention that (1) is simply a rewriting of (6). Furthermore, an astute reader, encountering that sentence will realize that the scope of the technical contribution of this paper is broad because the algorithmic framework developed here will be applicable to several other non-power system applications (mentioned in the two cited references in that sentence) where the second order Kuramoto model is used. Since this is just a matter of one sentence, we will prefer to keep it. We hope you approve.

\ul{On the other hand, the detailed discussion regarding Wasserstein gradient flow and why that standard development cannot be applied in this case, is in fact, the heart of this paper}. \ul{A key contribution of this paper is to help distill out and tackle the core mathematical difficulty that hinders simply importing standard techniques from the optimal transport literature. To make the presentation tighter, in the revised manuscript, we have eschewed the Appendix and accordingly adjusted the material in Sec. III-D.}}}

### 2022-02-28 Papers/Stochastic_Uncertainty_Propagation_TPS/RebuttalRound1-TPWRS-01391-2021.tex

\noindent
{\bf R2.2. Comments by reviewer}:\\
{\em The paper's language and presentation make it seem like it is more oriented towards the applied mathematics and control communities. It is highly suggested that the authors revise the exposition of ideas and concepts, putting more emphasis on the power system so that it can be well-understood by the power system experts.}

{\nib \blue{We appreciate the comment. \ul{In the revised manuscript, we have made some changes toward this direction. This includes slightly expanded overview of the related power system literature (see the two sentences added in Sec. I.1, second paragraph with two new references [22], [23]), citing recent reference [28] in Sec. I.1, third paragraph. Please note that the literature review in Sec. I.1 is completely based on the power system literature}.

Having said that, we believe our technical contribution is indeed interdisciplinary in the sense it is not mere translational research. In other words, we are not simply importing ideas from a different domain and applying them to a power system problem. Instead, we make novel contributions at the (possibly unanticipated) intersection of two areas: measure-valued gradient flows and power system dynamics. The presentation of the paper reflects the same, perhaps rightfully so. Notice in particular that our presentation is bottom-up: we start from power system literature review and mention that the problem addressed in our paper was already recognized as a technical challenge in that literature. We move onto the theoretical ideas and tools much later in the manuscript, as the build up demands.

We expect a significant share of the readers of this work to be power system engineers, but not exclusively so since this contribution will also be of interest to the researchers in control, optimization and applied mathematics. We do not believe this issue is exceptional for our paper since several high impact papers in the IEEE Trans. Power Systems were indeed at the intersection of these areas, and have positively contributed toward increasing interdisciplinary research collaboration.}}

@@

{\nib \blue{In this paper, we are interested in propagating the joint PDF values of the states of a power system dynamic model subject to joint stochastic uncertainties. Our focus is on explicitly computing these transient joint state PDFs at the time-varying state vector samples, which amounts to co-evolving the states and the joint PDF values over time. \ul{This is different from Monte Carlo or the traditional density estimation techniques since they approximate the joint PDFs via post-processing, not as an online computation}. To the best of our knowledge, there does not exist a power system simulation tool which can do this online computation while making neither statistical nor dynamical approximation.

Please note that in Sec. V, we have used MATPOWER and the Python-library ANDES (with inline citation of refs. [62] and [63] therein) for obtaining some simulation parameters, as explained in the first paragraph of V-A.}}

\noindent
{\bf R2.5. Comments by reviewer}:\\
{\em With regard to the point above, this reviewer would expect to see uncertainty bounds for the system dynamic response such as the trajectory of the rotor angle. Without such information, the method's usability is difficult to gauge.}

{\nib \blue{Thanks for this comment. \ul{The revised manuscript's newly incorporated Fig. 8 shows the statistical uncertainty bounds for the rotor angle sample paths for the simulation in Sec. V-A}. The inline text citing this figure in Sec. V-A is highlighted in blue.}}

@@

\noindent
{\bf R4.1. Comments by reviewer}:\\
{\em The main contribution of this paper lies in applying results on the governing partial differential equation (PDE) for the joint PDF to avoid discretization of the state space, thereby circumventing computational challenges. The reviewer appreciates that this paper is rigorous in development and thinks it deals with an important problem.}

{\nib \blue{We appreciate the positive remark.}}

\noindent
{\bf R4.2. Comments by reviewer}:\\
{\em The key contribution of this paper is that the proposed method can reduce the computation burden when solving for the transient PDF. However, no comparison is provided with existing methods. The authors should provide comparisons to the exiting approaches, for example, based on finite element discretization or other numerical methods. The test example could be performed using the 5-generator system or larger systems. Besides, referring to Figs. 8 and 9, the total computation time should be reported for a longer horizon.}

{\nib \blue{\ul{Regarding the comparison with existing methods}, we have provided comparison with Monte Carlo in the revised manuscript's Fig. 11 {\red{and another?? (in response to comment \textbf{R4.7})}}. The new Fig. 8 in the revised manuscript also shows the consistency for the rotor angle stochastic predictions w.r.t. Monte Carlo for the simulation reported in Sec. V-A, as requested by comment \textbf{R2.5}. Please note that performing finite element simulation for 5-generator system is a very tall order since the state space then is 10 dimensional, and even if we do a very coarse 10 discretizations per dimension, then such a simulation would require us to evolve a joint PDF over $10^{10}$ grid points. This requires dedicated high-performance computing architecture and an undertaking well beyond the scope of this paper. This also explains why the state-of-the-art for direct computation of transient joint PDF in power systems literature remains single-machine-infinite-bus $(n=1)$ as in the 2013 ref. [45] which indeed performed the finite element simulation with a two dimensional discretization, and mentioned the difficulty of scaling this computational approach for even moderately large sized systems.

\ul{Regarding the computational time for longer horizon}, {\red{TBD}}}}

### 2022-02-28 Papers/Stochastic_Uncertainty_Propagation_TPS/JKO_Power_Systems_TPS_Rev1.tex
polishing rebuttal round 1

@@

Fig. \ref{fig:IEEE14omegamarginalsCase12} shows the corresponding univariate $\omega_{i}$-marginal PDFs. In particular, Fig. \ref{fig:IEEE14omegamarginalsCase12}(c) helps compare the $\omega_{4}$ marginal evolution for bus 6 in the nominal (\textbf{Case I}) and post-fault (\textbf{Case II}) scenarios. Since \textbf{Case II} considers the failure of line 13, from Fig. \ref{fig:Motivating} we intuitively expect that the transient effects of this fault will be prominent for bus 6 (generator 4). Indeed, Fig. \ref{fig:IEEE14omegamarginalsCase12}(c) shows that although the $\omega_{4}$ marginals in both \textbf{Case I} and \textbf{II}, at all times have the highest probability around $0$ rad/s, the post-fault marginals show larger spread or dispersion of the probability mass, as expected. That the generator 4 states in \textbf{Case II} have larger uncertainties than in \textbf{Case I}, is also visible in the bivariate marginal plots if we compare the fourth row in Fig. \ref{fig:IEEE14marginalsCase0} with the same in Fig. \ref{fig:IEEE14marginalsCase1}.

>>>>

\begin{figure*}[ht!]
   \subfigure[Case I]{\includegraphics[width=0.49\textwidth]{IEEE14_Case0_RotorAngles.png}}
   \hspace{\fill}
   \subfigure[Case II]{\includegraphics[width=0.49\textwidth]{IEEE14_Case1_RotorAngles.png}}
\caption{\small{Boxplots for the rotor angles $\theta_{i}(t)$, $i=1,\hdots,5$, for the IEEE 14 bus system simulation in Sec. \ref{subsec:NumSimIEEEBus} in the (a) nominal \textbf{Case I}, and in the post-fault \textbf{Case II} scenarios, respectively. As usual, the boxes show the respective interquartile ranges, and the whiskers show the respective minimum and maximum values. In each box plot, we have superimposed  the respective \emph{mean} rotor angle sample path.}}
\vspace*{-0.1in}
\label{fig:IEEE14theta}
\end{figure*}

@@

Fig. \ref{fig:IEEE14omegamarginalsCase12} shows the corresponding univariate $\omega_{i}$-marginal PDFs. In particular, Fig. \ref{fig:IEEE14omegamarginalsCase12}(c) helps compare the $\omega_{4}$ marginal evolution for bus 6 in the nominal (\textbf{Case I}) and post-fault (\textbf{Case II}) scenarios. Since \textbf{Case II} considers the failure of line 13, from Fig. \ref{fig:Motivating} we intuitively expect that the transient effects of this fault will be prominent for bus 6 (generator 4). Indeed, Fig. \ref{fig:IEEE14omegamarginalsCase12}(c) shows that although the $\omega_{4}$ marginals in both \textbf{Case I} and \textbf{II}, at all times have the highest probability around $0$ rad/s, the post-fault marginals show larger spread or dispersion of the probability mass, as expected. That the generator 4 states in \textbf{Case II} have larger uncertainties than in \textbf{Case I}, is also visible in the bivariate marginal plots if we compare the fourth row in Fig. \ref{fig:IEEE14marginalsCase0} with the same in Fig. \ref{fig:IEEE14marginalsCase1}. {\blue{The statistical uncertainties in the rotor angles are shown in Fig. \ref{fig:IEEE14theta} for both \textbf{Case I} and \textbf{Case II}.}}
